Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-04-27 paper.tex
revision some chages part1

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studies structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle
triangle pattern. This pattern is much easier to vectorize and parallelize
compared to previously used random triangle or spider-web sampling
patterns. Moreover, it allows to easily implement computations of directional
3-point CFs. The execution times for the same number of sample are orders of
magnitude lower than existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun(accepted)]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis. Examples of such objects
for studies include galaxy formations \cite{springel2006}, immiscible multiphase
fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples
\cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010},
food specimens \cite{derossi2019,nagdalian2021}, biological tissue
\cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and
interstellar turbulence \cite{portillo2018developing}. The scale of interest may
span orders of magnitude from angstroms and nm for materials
\cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for
star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can
be non-static -- it will change with time, and this dynamics is usually also of
great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure
is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of
    theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous
    bounds and such;
  \item Prediction of material physical properties using machine (deep) learning
    \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison
    and analysis
    \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,
      portillo2018developing,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular},
    which may include temporal dynamics or surface evolution
    \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity
    \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity
    analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data
    \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023}
    in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature
    extraction), and retrieval (by means of stochastic reconstructions)
    \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off
    between field-of-view and imaging resolution
    \cite{SciRep1,Geoderma2018,chen2020super,karimpouli2022}.
\end{enumerate}

@@

However, while 2-point CFs are computed and used in multitude of studies, higher
order functions, for example, 3-point correlation functions are not that widely
used. The reason for this is straightforward -- 3-point correlation functions
are more involved and, thus, require much more computational resources to
assess. It has been shown that the amount of additional information content with
each increase in number of points $n$ decreases \cite{yao1993high,Gommes2}. The
trade-off between additional information and computational burden is still not
clear. Nonetheless, the computation of 3-point statistics is known to be
advantageous at least in the number of applications as related to cosmology,
astrophysics and turbulence patterns analysis as they can probe non-Gaussian
signatures in the distribution of matter and energy, which are not captured by their
2-point counterparts
\cite{TakadaJain,hopkins2013stars,gorbunova2016precessing,yoo2022non}.

>>>>

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studied structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle
triangle pattern. This pattern is much easier to vectorize and parallelize
compared to previously used random triangle or spider-web sampling
patterns. Moreover, it allows to easily implement computations of directional
3-point CFs. The execution times for the same number of samples are orders of
magnitude lower than in existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun., 299, 109134]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis. Examples of such objects
for studies include galaxy formations \cite{springel2006}, immiscible multiphase
fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples
\cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010},
food specimens \cite{derossi2019,nagdalian2021}, biological tissue
\cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and
interstellar turbulence \cite{portillo2018developing}. The scale of interest may
span orders of magnitude from angstroms and nm for materials
\cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for
star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can
be non-static -- it will change with time, and this dynamics is usually also of
great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure
is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of
    theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous
    bounds and such;
  \item Prediction of material physical properties using machine (deep) learning
    \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison
    and analysis
    \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,
      portillo2018developing,cheng2022data,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular},
    which may include temporal dynamics or surface evolution
    \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity
    \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity
    analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data
    \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023}
    in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature
    extraction), and retrieval (by means of stochastic reconstructions)
    \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off
    between field-of-view and imaging resolution
    \cite{SciRep1,Geoderma2018,chen2020super,yan2023multiscale,karimpouli2022,yan2024modeling}.
\end{enumerate}

@@

However, while 2-point CFs are computed and used in multitude of studies, higher
order functions, for example, 3-point correlation functions are not that widely
used. The reason for this is straightforward -- 3-point correlation functions
are more involved and, thus, require much more computational resources to
assess. It has been shown that the amount of additional information content decreases with
each increase in number of points $n$ \cite{yao1993high,Gommes2}. The
trade-off between additional information and computational burden is still not
clear. Nonetheless, the computation of 3-point statistics is known to be
advantageous at least in the number of applications as related to cosmology,
astrophysics and turbulence patterns analysis as they can probe non-Gaussian
signatures in the distribution of matter and energy, which are not captured by their
2-point counterparts
\cite{TakadaJain,hopkins2013stars,gorbunova2016precessing,yoo2022non}.
